|

### 2025-02-21 output/Main_text.Rmd

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
the R package *psych* [@revelle_psych_2024], graphically represents
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970]:
Indeed, the cosine between any two axes (or, more generally, vectors)
is considered as the geometrical equivalent of the correlation between
the variables they represent.
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
intends to model responses to certain stimuli and estimate person scores,
being the model interpretation secondary in relevance.
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constraints on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
critical misconceptions about the measurement model.

Take the following as an example:
A hypothetical complex item measures two latent traits with equal validity
---i.e., its discrimination parameters are equal with value $`r DISCR_SYM`$.
Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
we would find that $`r AG_MDISC_EXAMPLE`$.
Assume we apply this item in two groups where, taken to the extreme,
the two measured dimensions are orthogonal in the first group
and perfectly correlated in the second one.
While the previous formula seems correct in the first group,
one can intuitively say that,
because the two dimensions are *perfectly aligned* and thus
compound up in the second group,
the item `r MDISC_SYM` should be the algebraic sum of both components,
that is, $`r MDISC_VALUE_COLLIN`$.
Similarly, the correlation between two simple items items,
each one measuring a distinct dimension,
should be uncorrelated in the first group,
but perfectly correlated in the second one.
However, using an orthogonal representation
would yield null cosines in both cases.

Therefore, the representation of a latent space where the traits are correlated
is more accurate on a geometrical coordinate system with non-orthogonal axes
[@ackerman_multidimensional_2005-1, p. 16, see footnote 1]
and (possibly) non-standard units,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g., multigroup equating) even when
interpreting the latent dimensions as substantive constructs is not a goal
[for multidimensional linking and equating procedures,
see Chapter 9 in @reckase_multidimensional_2009].
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

On an unrelated basis, negative loadings/discrimination parameters
are often irrelevant in MIRT applications in the cognitive testing domain.
It has been argued that discrimination parameters
"are constrained to be positive" [@ackerman_graphical_1996, p. 315],
or that "in an [independent-clusters] solution, the loadings can always
be selected to be positive" [@mcdonald_basis_2000, p. 110].
However, when such instances of negative parameters appear,
they are often dismissed as "particularly puzzling" (p. 109),
or implicitly assuming item malfunctioning without further discussion
[e.g., @ackerman_multidimensional_2005-1].
In the non-cognitive domain though, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
usually reverse-scored--- is widespread
[consider, e.g., the statement "Neglect my duties" in the Conscientiousness item
of the Big-Five Factor Markers, @noauthor_international_nodate].
Instruments that combine positively- and negatively-keyed items
are relatively common, and
complex indicators may even have discrimination parameters
with opposite signs for different latent dimensions
[see, e.g., @mclarnon_be_2016,
for item parcels with both positive and negative loadings,
on method factors in the case of the negative ones].
Following @ackerman_multidimensional_2005-1, one can easily note that
a monotonically decreasing probability for a certain dimension
is given by a non-positive discrimination parameter
(and an unchanging probability by a null discrimination parameter,
which also applies to cognitive instruments),
contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411]
for determining the multidimensional difficulty.
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim to provide a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
To pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application,
which will require introducing some linear algebra concepts and results.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce three different versions,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss their properties,
propose a graphical representation method based on the new formulation,
and present two examples of application of our results.
These examples, from the cognitive and the non-cognitive domain respectively,
are taken from the MIRT literature:
One of them is a classic dataset, often used to exemplify
the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
the other one, an application in marketing research [@tezza_modelo_2018].
We round up with a discussion about
the applicability and generalization of our results.

@@

Consider a basis $`r LS_BASIS_EQ`$,
with $`r LS_BASIS`$ generally non-orthonormal.
Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
in basis $`r LS_BASIS`$ (with $`r N_DIMS`$ any positive integer),
where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
According to the `r MODEL_ACRONYM` model, the probability of
a positive response to item $`r ITEM_INDEX`$ is

@@

We must stop here and make a few remarks. First, as we are not constraining
ourselves to cognitive traits,
we speak of _positive_ instead of _correct response_. This may imply giving
a correct response indeed, if we are talking about a cognitive test.
But it may mean endorsing an item (or a certain response option)
in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in @reckase_difficulty_1985
---a decrease in the probability of a correct response
with an increase in ability makes little sense.
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
Third, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
instead of the item difficulty.
Finally, it is worth highlighting that
this model generally assumes *within-item multidimensionality*,
which will be present whenever more than one component of $`r DISCR_VECTOR`$
is non-null.
However, the model can also account for *between-item multidimensionality*,
if the $`r DISCR_VECTOR`$ parameter only has one non-null component.

@@

Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
with the dot product as inner product.
Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_ORTH_BASIS`$,
respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
the linear combination of its element vectors
scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.

@@

Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
express the item coordinates in terms of their multidimensional parameters.
However, it is worth noting that the latent space Gram matrix
also plays a role in computing these coordinates,
and will therefore have an effect on the representation.
Of course,
from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
it is easy to express the coordinates in terms of the parameters in the original
model formulation.
Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forward or backward with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.
Of course, it is also directly proportional to
the intercept parameter $`r INTERCEPT_PARAM`$.
This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
regardless of the discrimination.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version, however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

The latter effect can be better noticed in another case, in which Reckase
uses the same item parameters to demonstrate the over-specification of
dimensions (see p. 183).
In that example, he uses a covariance matrix that practically has rank 1,
implying that the (alleged) three dimensions collapse into a single one, being
thus the test unidimensional in practice.
In such a case, the covariance-based direction angles
are extremely close to 0 for all items, as one would expect
from a unidimensional test, where all items are strictly aligned with the one
dimension they measure. The reader interested can check out the parameters
computed in that case in the table in Online Resource 1.

@@

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameters have the drawback of not fulfilling
the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
namely, the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is relevant in applications where
potentially different bases are involved.
Furthermore, not that, when using the covariance-based version formulation,
the resulting norm in the latent space is the Mahalanobis distance
[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
thus taking the latent covariance structure into account. 
Given these considerations,
we propose adopting the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad-hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution and apply
an oblique rotation afterwards, at least in exploratory solutions.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution applying
(the covariance-based version of) Equation \@ref(eq:dir-cos-vec-test-space).
<!-- TODO: Number and reference cov-based version of the cosines? -->
What a researcher should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
instead of the invariant value of $`r MDISC_COV_PARAM`$.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
(Note that in the original derivations by @reckase_difficulty_1985
and @reckase_discriminating_1991 both matrices simplify to the identity matrix,
so there was no need to make this distinction.)
This implies that the axes defined by the test space basis,
and not the ones defined by the latent space basis,
are a geometrical representation of the covariance structure:
Indeed, the cosines of the angles among the test space axes are
the geometric equivalent of the correlations.
This result is both surprising and counterintuitive
---Several instances in the MIRT literature suggest that
"the angle between the axes [of the latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
We hope our results help shed a clearer light on
this often misunderstood topic.

>>>>

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
the R package *psych* [@revelle_psych_2024], graphically represents
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
intends to model responses to certain stimuli and estimate person scores,
being the model interpretation secondary in relevance.
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constraints on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
critical misconceptions about the measurement model.

The representation of a latent space where the traits are correlated is more
accurate on a geometrical coordinate system with non-orthogonal axes
[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g., multigroup equating) even when
interpreting the latent dimensions as substantive constructs is not a goal
[for multidimensional linking and equating procedures,
see Chapter 9 in @reckase_multidimensional_2009].
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

On an unrelated basis, negative loadings/discrimination parameters
are often irrelevant in MIRT applications in the cognitive testing domain.
It has been argued that discrimination parameters
"are constrained to be positive" [@ackerman_graphical_1996, p. 315],
or that “in an [independent-clusters] solution, the loadings can always
be selected to be positive” [@mcdonald_basis_2000, p. 110].
However, when such instances of negative parameters appear,
they are often dismissed as “particularly puzzling” (p. 109),
or implicitly assuming item malfunctioning without further discussion
[e.g., @ackerman_multidimensional_2005-1].
In the non-cognitive domain though, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
usually reverse-scored--- is widespread
[consider, e.g., the statement "Neglect my duties" in the Conscientiousness item
of the Big-Five Factor Markers, @noauthor_international_nodate].
Instruments that combine positively- and negatively-keyed items
are relatively common, and
complex indicators may even have discrimination parameters
with opposite signs for different latent dimensions
[see, e.g., @mclarnon_be_2016,
for item parcels with both positive and negative loadings,
on method factors in the case of the negative ones].
Following @ackerman_multidimensional_2005-1, one can easily note that
a monotonically decreasing probability for a certain dimension
is given by a non-positive discrimination parameter
(and an unchanging probability by a null discrimination parameter,
which also applies to cognitive instruments),
contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411]
for determining the multidimensional difficulty.
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim to provide a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
To pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application,
which will require introducing some linear algebra concepts and results.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce three different versions,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss their properties,
propose a graphical representation method based on the new formulation,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
in basis $`r LS_BASIS_EQ`$, being $`r LS_BASIS`$ not necessarily orthonormal,
where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
According to the `r MODEL_ACRONYM` model, the probability of
a positive response to item $`r ITEM_INDEX`$ is

@@

We must stop here and make a few remarks. First, as we are not constraining
ourselves to cognitive traits,
we speak of _positive_ instead of _correct response_. This may imply giving
a correct response indeed, if we are talking about a cognitive test.
But it may mean endorsing an item (or a certain response option)
in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in @reckase_difficulty_1985
---a decrease in the probability of a correct response
with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
instead of the item difficulty.

@@

Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
with the dot product as inner product.
Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_ORTH_BASIS`$,
respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.

@@

Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
express the items coordinates in terms of their multidimensional parameters.
However, it is worth noting that the latent space Gram matrix
also plays a role in computing these coordinates,
and will therefore have an effect on the representation.
Of course,
from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
it is easy to express the coordinates in terms of the parameters in the original
model formulation, namely,

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Hence we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

The latter effect can be better noticed in another case, in which Reckase
uses the same item parameters to demonstrate the over-specification of
dimensions (see p. 183).
In that example, he uses a covariance matrix that practically has rank 1,
implying that the (alleged) three dimensions collapse into a single one, being
thus the test practically unidimensional.
In such a case, the covariance-based direction angles
are extremely close to 0 for all items, as one would expect
from a unidimensional test, where all items are strictly aligned with the one
dimension they measure. The reader interested can check out the parameters
computed in that case in the table in Online Resource 1.

@@

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameters have the drawback of not fulfilling
the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
namely, the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is important in applications where
potentially different bases are involved.
Given these considerations,
we propose to adopt the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad-hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require computing the cosines in
the original basis afterwards.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
This implies that the axes defined by the test space basis,
and not the latent space basis,
are a geometrical representation of the covariance structure.
This result is both surprising and counterintuitive
---Several instances in the MIRT literature suggest that
"the angle between the axes [of the latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
We hope our results help shed a clearer light on
this often misunderstood topic.

### 2025-02-21 output/sandbox_slideshow.qmd

---
bibliography:  ../www/Multidimensional-parameters-MCLM.bib
csl:           ../www/apa-old-doi-prefix.csl
date-meta:     "2024-11-27"
editor:
  mode:        source
  markdown: 
    wrap:      sentence
    canonical: true
lang:          es
knitr:
  opts_chunk: 
    results:   asis
    cache:     false
    autodep:   true
format:
  revealjs:
    auto-stretch:            true
    code-annotations:        hover
    df-print:                paged
    fig-cap-location:        bottom
    fig-format:              svg
    fig-asp:                 1
    fig-width:               6
    keep-md:                 true
    link-external-newwindow: true
    margin:                  0
    self-contained:          true
    slide-number:            false
    theme:                   ../www/slideshow-assets/extra-styles.scss
editor_options: 
  chunk_output_type: console
---

@@

```{r pre}
PROB_AXIS    <- seq(-3, 3, by = .01)
ERROR_POINTS <- seq(-3, 3, length.out = 5)
PROB_AXIS_2D <- seq(-3, 3, by = .1)
F1_COLOR <- unname(PALETTE_BLUE["normal"])

@@

ITEMS_PALETTE <- c(
  PALETTE_BLUE["normal"],
  PALETTE_RASPBERRY["normal"],
  PALETTE_APPLE["normal"],
  PALETTE_TANGERINE["normal"]
) |>
  unname() |>
  setNames(paste(1:4, sep = '_'))
```

>>>>

---
title:     |
  Sandbox slide
bibliography:  ../www/Multidimensional-parameters-MCLM.bib
csl:           ../www/apa-old-doi-prefix.csl
date:          "2024-10-21"
date-meta:     "2024-11-27"
date-format:   long
editor:
  mode:        source
  markdown: 
    wrap:      sentence
    canonical: true
lang:          es
knitr:
  # opts_knit: # Does not work, see line 88
  #   root_dir:  here::here()
  opts_chunk: 
    results:   asis
    cache:     false
    autodep:   true
format:
  revealjs:
    auto-stretch:            true
    code-annotations:        hover
    df-print:                paged
    fig-cap-location:        bottom
    fig-format:              svg
    fig-asp:                 1
    fig-width:               6
    incremental:             false
    keep-md:                 true
    link-external-newwindow: true
    margin:                  0
    self-contained:          true
    slide-number:            false
    theme:                   ../www/slideshow-assets/extra-styles.scss
    transition:              none
    view-distance:           3
    template-partials:
      - ../www/slideshow-assets/title-slide.html
editor_options: 
  chunk_output_type: console
---

@@

```{r pre}
PROB_AXIS_2D <- seq(-3, 3, by = .1)
F1_COLOR <- unname(PALETTE_BLUE["normal"])

@@

```
